---
title: "Ch 10 Bootstrap"
output: html_notebook
---

# Example: median income

Suppose the following income distribution:

```{r}
set.seed(1995)
n <- 10^6
income <- 10^(rnorm(n, log10(45000), log10(3)))
hist(income/10^3, nclass = 1000)
```

The population median is

```{r}
m <- median(income)
m
```

Estimate the median from a sample of 100 estimates:

```{r}
N <- 100
x <- sample(income, N)
median(x)
```

# Confidence intervals for the median

Because we are simulating the data, we can use a Monte Carlo simulation to learn the distribution of $M$.

```{r}
library(gridExtra)
B <- 10^4
m <- replicate(B, {
  x <- sample(income, N)
  median(x)
})
```

```{r median-is-normal, message = FALSE, warning = FALSE, out.width="100%", fig.width = 6, fig.height = 3}
rafalib::mypar(1,2)
hist(m, nclass = 30)
qqnorm(scale(m));abline(0,1)
```

The CLT applies to averages, but here we are interested in the median. The confidence interval based on CLT would be

```{r}
median(x) + 1.96*sd(x)/sqrt(N)*c(-1,1)
```

Which is far from the actual

```{r}
quantile(m, c(0.025, 0.975))
```

So a series of "bootstrap samples" can be used to get closer.

```{r}
B <- 10^4
m_star <- replicate(B, {
  x_star <- sample(x, N, replace = TRUE)
  median(x_star)
})
```

```{r}
quantile(m_star, c(0.025, 0.975))
```
