---
title: "Ch 21 Applied Linear Algebra"
output: html_notebook
---

```{r}
library(tidyverse)
library(dslabs)
if (!exists("mnist")) mnist <- read_mnist()
x <- mnist$train$images 
y <- mnist$train$labels
```

# Matrix multiplication

```         
m <- nrow(A)
p <- ncol(B)
C <- matrix(0, m, p)
for(i in 1:m){
  for(j in 1:p){
    C[i,j] <- sum(A[i,] * B[,j])
  }
}
```

In R, use `%*%`:

```         
C <- A %*% B
```

$$
\mathbf{A}\mathbf{x} =  \mathbf{b}
$$

$$
\mathbf{A}^{-1}\mathbf{A}\mathbf{x} =   \mathbf{x} = \mathbf{A}^{-1} \mathbf{b}
$$

To solve the first equation

```{r}
A <- matrix(c(1, 3, -2, 3, 5, 6, 2, 4, 3), 3, 3, byrow = TRUE)
b <- matrix(c(5, 7, 8))
solve(A, b)
```

# Identity matrix

Because the default for the second argument in `solve` is an identity matrix, if we simply type `solve(A)`, we obtain the inverse $\text{A}^{-1}$. This means we can also obtain a solution to our system of equations with:

```{r}
solve(A) %*% b
```

```{r}
A
solve(A)
```

# Distance

For $n$ features the distance is

$$
|| \mathbf{x}_1 - \mathbf{x}_2 || = \sqrt{ \sum_{j=1}^{n} (x_{1,j}-x_{2,j })^2 }
$$

```{r}
x_1 <- x[6,]
x_2 <- x[17,]
x_3 <- x[16,]
```

Computing distance between each pair

```{r}
c(sum((x_1 - x_2)^2), 
  sum((x_1 - x_3)^2), 
  sum((x_2 - x_3)^2)) |> sqrt()
```

```{r}
c(crossprod(x_1 - x_2), 
  crossprod(x_1 - x_3), 
  crossprod(x_2 - x_3)) |> sqrt()
```

```{r}
y[c(6, 17, 16)]
```

We can also compute **all** the distances at once relatively quickly using the function `dist`, which computes the distance between each row and produces an object of class `dist`:

```{r}
d <- dist(x[c(6, 17, 16),])
class(d)
```

```{r}
d
```

Compute the distance between each of the first 300 observations and then make an image:

```{r fig.width = 5, fig.height = 5}
d <- dist(x[1:300,])
image(as.matrix(d))
```
