---
title: "Continuous Probability"
output: html_notebook
---

# Continuous Probability

```{r}
library(tidyverse)
library(dslabs)
library(plotly)
```

```{r}
x <- heights %>% filter(sex == "Male") %>% 
  pull(height)
```

# eCDF and Cumulative Distribution Function

eDF is defined:

```{r}
F <- function(a) mean(x <= a)
```

What is the chance the man is taller than 70.5 inches:
```{r}
1 - F(70)
```

CDF:

$$F(a) = \mbox{Pr}(X \leq a)$$
Probability of a student being between height $a$ and $b$:

$$\mbox{Pr}(a < X \leq b) = F(b)-F(a)$$

## Probability Density Function

$$F(b) - F(a) = \int_a^b f(x)\,dx$$

Gamma distribution

```{r}
cont <- data.frame(x = seq(0, 5, len = 300), y = dgamma(seq(0, 5, len = 300), 2, 2))
disc <- data.frame(x = seq(0, 5, 0.075), y = dgamma(seq(0, 5, 0.075), 2, 2))
ggplot(mapping = aes(x, y)) +
  geom_col(data =  disc) +
  geom_line(data = cont) +
  ylab("f(x)") -> p
ggplotly(p)

```

PDF for normal distribution:

$$f(x) = e^{-\frac{1}{2}\left( \frac{x-m}{s} \right)^2}$$

```{r, eval=FALSE}
F(a) = pnorm(a, m, s)
```

```{r}
m <- mean(x)
s <- sd(x)
1 - pnorm(70.5, m, s)
```

```{r}
mean(x <= 68.5) - mean(x <= 67.5)
pnorm(68.5, m, s) - pnorm(67.5, m, s) 
mean(x <= 69.5) - mean(x <= 68.5)
pnorm(69.5, m, s) - pnorm(68.5, m, s) 
mean(x <= 70.5) - mean(x <= 69.5)
pnorm(70.5, m, s) - pnorm(69.5, m, s) 
```

**But** it doesn't work if the interval is too small.

```{r}
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, m, s) - pnorm(70.1, m, s)
mean(x <= 68.9) - mean(x <= 68.1)
pnorm(68.9, m, s) - pnorm(68.1, m, s)
```

```{r}
mean(x <= 69.9) - mean(x <= 66.1)
pnorm(69.9, m, s) - pnorm(66.1, m, s)
```

# The Probability Density

Use the normal distribution to estimate probability taller than 76 in

```{r}
1 - pnorm(76, m, s)
```

```{r}
dat <- tibble(x = seq(-4, 4, length = 100)*s + m,
              y = dnorm(x, m, s))

dat_ribbon <- filter(dat, x >= 2*s + m)

ggplot(dat, aes(x, y)) +
  geom_line() +
  geom_ribbon(aes(ymin = 0, ymax = y), data = dat_ribbon) -> p
ggplotly(p)
```

# Monte Carlo

```{r}
n <- length(x)
m <- mean(x)
s <- sd(x)
simulated_heights <- rnorm(n, m, s)
```

```{r}
data.frame(simulated_heights = simulated_heights) |>
  ggplot(aes(simulated_heights)) + 
  geom_histogram(color = "black", binwidth = 1) 
```

If, for example, we pick 800 males at random, what is the distribution of the tallest person? How rare is a seven-footer in a group of 800 males? 

```{r}
B <- 10000
tallest <- replicate(B, {
  simulated_data <- rnorm(800, m, s)
  max(simulated_data)
})
```

```{r}
mean(tallest >= 7*12)
```

Such a person is quite rare.

```{r}
data.frame(tallest = tallest) %>% 
  ggplot(aes(tallest)) +
  geom_histogram(color='black', binwidth = 1)
  
```

The distribution does not look normal.

# Continuous distributions


```{r}
data.frame(x, f = dnorm(x)) %>% 
  ggplot(aes(x, f)) +
  geom_line()
```
















