---
title: "Discrete Probability"
output: html_notebook
---

# Discrete probability

```{r}
set.seed(1986)
```

## Combinations and permutations

```{r}
letters[1:5]
```

### Deck of cards

```{r}
suits <- c("of Diamonds", "of Clubs", "of Hearts", "of Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six",
             "Seven", "Eight", "Nine", "Ten", "Jack", "Queen",
             "King")
deck <- expand.grid(number = numbers, suit=suits)
deck <- paste(deck$number, deck$suit)
deck

```

### Deck of cards

```{r}
suits <- c("of Diamonds", "of Clubs", "of Hearts", "of Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six",
             "Seven", "Eight", "Nine", "Ten", "Jack", "Queen",
             "King")
deck <- expand.grid(number = numbers, suit=suits)
deck <- paste(deck$number, deck$suit)
deck

```

```{r}
kings <- paste("King", suits)
kings
```

```{r}
mean(deck %in% kings)
```

```{r}
library(gtools)
```

```{r}
permutations(10,7) |> head(10)
```

With a size vector

```{r}
permutations(10,7, v = 10:19) |> head(10)
```

```{r}
all_phone_numbers <- permutations(10, 7, v = 0:9)
n <- nrow(all_phone_numbers)
index <- sample(n, 5)
all_phone_numbers[index,]
```

## All possible hands

With two cards

```{r}
hands <- permutations(52, 2, v = deck)
head(hands, 6)
```

```{r}
hands[2:4,2]
```

```{r}
first_card <- hands[,1]
second_card <- hands[,2]
```

```{r}
sum(first_card %in% kings)
```

## Conditional probability that second card is also a king

$$\mbox{Pr}(B \mid A) = \frac{\mbox{Pr}(A \mbox{ and } B)}{ \mbox{Pr}(A)}$$

```{r}
print(mean(first_card %in% kings & second_card %in% kings) /
  mean(first_card %in% kings))
```

Order doesn't matter: chance of natural 21

```{r}
combinations(3,2)
```

```{r}
aces <- paste("Ace", suits)
facecard <- c("King", "Queen", "Jack", "Ten")
facecard <- expand.grid(number = facecard, suit = suits)
facecard <- paste(facecard$number, facecard$suit)
```

```{r}
hands <- combinations(52, 2, v = deck)
mean((hands[,1] %in% aces & hands[,2] %in% facecard) |
       (hands[,2] %in% aces & hands[,1] %in% facecard))
```

### Monte Carlo example

```{r}
hand <- sample(deck, 2)
hand
```

```{r}
(hand[1] %in% aces & hand[2] %in% facecard) | 
  (hand[2] %in% aces & hand[1] %in% facecard)
```

```{r}
blackjack <- function(){
  hand <- sample(deck, 2)
  (hand[1] %in% aces & hand[2] %in% facecard) | 
  (hand[2] %in% aces & hand[1] %in% facecard)
}
```

```{r}
blackjack()
```

```{r}
B <- 10000
results <- replicate(B, blackjack())
mean(results)
```

## Monty Hall Problem

```{r}
B <- 10000
monty_hall <- function(strategy){
  doors <- as.character(1:3)
  prizes <- c("car", "goat", "goat")
  # called w/o second argument, sample returns all values
  # in random order
  prize <- sample(prizes)
  prize_door <- doors[prize == "car"]
  my_pick <- sample(doors, 1)
  show <- sample(doors[!doors %in% c(my_pick, prize_door)],1)
  stick <- my_pick
  stick == prize_door
  switch <- doors[!doors %in% c(my_pick, show)]
  choice <- ifelse(strategy == "stick", stick, switch)
  choice == prize_door
}
```

```{r}
stick <- replicate(B, monty_hall("stick"))
mean(stick)
```

```{r}
switch <- replicate(B, monty_hall("switch"))
mean(switch)
```

## Birthday problem

Group of 50 people, probability that at least 2 have the same birthday.

```{r}
n <- 50
bdays <- sample(1:365, n, replace = TRUE)
```

```{r}
duplicated(bdays)
```

```{r}
any(duplicated(bdays))
```

```{r}
B <- 10000
same_birthday <- function(n){
  bdays <- sample(1:365, n, replace = TRUE)
  any(duplicated(bdays))
}
results <- replicate(B, same_birthday(n))
mean(results)
```

```{r}
compute_prob <- function(n, B = 10000){
  results <- replicate(B, same_birthday(n))
  mean(results)
}
```

```{r}
library(tidyverse)
n <- seq(1,60)
prob <- sapply(n, compute_prob)
qplot(n, prob)
```

```{r}
data.frame(n, prob) |>
  ggplot(aes(n, prob)) +
  geom_point()

```

Compute the actual probabilities

$$1 \times \frac{364}{365}\times\frac{363}{365} \dots \frac{365-n + 1}{365}$$

```{r}
exact_prob <- function(n){
  prob_unique <- seq(365, 365 - n + 1) / 365
  1 - prod(prob_unique)
}
eprob <- sapply(n, exact_prob)
ggplot() +
  geom_point(aes(n, prob)) +
  geom_line(aes(n, eprob), col="red")
```

## Testing stability of estimate

```{r}
B <- 10^seq(1, 5, len = 100)
compute_prob <- function(B, n = 25){
  same_day <- replicate(B, same_birthday(n))
  mean(same_day)
}
prob <- sapply(B, compute_prob)
plot(log10(B), prob)
```

Values start to stabilize around 1000. At 10000 they are quite stable.
